---
title: "Class08"
author: "Alexander Liu (PID: 69026918)"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

Q1. How many different candy types are in this dataset?
```{r}
nrow(candy)
```
85 types.

Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity == 1)
```
38 types.

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["100 Grand",]$winpercent
```
My favorite is 100 Grand.
Its `winpercent` is 66.97173

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat",]$winpercent
```
Kit Kat's winpercent is 76.7686.

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
Tootsie Roll Snack Bars's winpercent is 49.6535.

```{r}
library("skimr")
skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent.

Q7. What do you think a zero and one represent for the candy$chocolate column?

"1" indicates that the variable is chocolate, whereas "0" indicates not.

Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```

Q9. Is the distribution of winpercent values symmetrical?

According to the histogram, not.

Q10. Is the center of the distribution above or below 50%?
```{r}
median(candy$winpercent)
```
Below 50%.

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
chocolate <- candy[candy$chocolate == 1, "winpercent"]
fruity <- candy[candy$fruity == 1, "winpercent"]
mean(chocolate)
mean(fruity)
```
Higher.

Q12. Is this difference statistically significant?
```{r}
t.test(chocolate, fruity)
```
Yes.

Q13. What are the five least liked candy types in this set?
```{r}
library(dplyr)
row.names.data.frame(head(candy[order(candy$winpercent),], n=5))
row.names.data.frame(candy %>% arrange(winpercent) %>% head(5))
```
"Nik L Nip", "Boston Baked Beans", "Chiclets", "Super Bubble", and "Jawbusters".

Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
row.names.data.frame(tail(candy[order(candy$winpercent),], n=5))
row.names.data.frame(candy %>% arrange(winpercent) %>% tail(5))
```
"Snickers", "Kit Kat", "Twix", "Reese's Miniatures", and "Reese's Peanut Butter cup".

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_bar(stat = "identity")
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

- Q17. What is the worst ranked chocolate candy?

 "Nik L Nip"

- Q18. What is the best ranked fruity candy?

"Reese's Peanut Butter cup"

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 50)
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

I calculated the cost efficiency (winpercent/pricepercent)

```{r}

candy$bang <- candy$winpercent/candy$pricepercent
ord <- order(candy$bang, decreasing = TRUE)
head( candy[ord,c(11,12,13)], n=5 )
```
"Tootsie Roll Midgies".



Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
tail( candy[ord,c(11,12,13)], n=5 )
```
"Nik L Nip".

Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col()
```

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
```{r}

```

chocolate vs fruity, and pluribus vs bar

Q23. Similarly, what two variables are most positively correlated?

chocolate vs bar, and chocolate vs winpercent

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,"PC1"], pca$x[,"PC2"])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 50)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
#| eval: false
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, and pluribus.
Many candies which have these characteristics can be found at the positive side of the PC1-PC2 plot.
